---
title: "DDfg_csLiteracy"
output: html_document
date: "2023-06-26"
---
Configuring Markdown file:
```{r setup}
knitr::opts_chunk$set(echo = FALSE, message = FALSE)
```

First, start by importing the necessary libraries:
```{r}
library(tidyverse)
library(ggplot2)
library(sf)
library(tmap)
library(readxl)

```

Importing data:
```{r}
# reading data on school info from https://profiles.doe.mass.edu/search/search.aspx?leftNavId=11238
# info gives us school address (town and zipcode)
schoolinfo <- read_excel("data/school_info.xlsx") %>% 
  select(`Org Name`, `Org Code`, `Org Type`, `Address 1`, Town, Zip) %>% 
  mutate(full_address = paste(`Address 1`, ", ", Town))

# trying two different sources (both too outdated so not used)
district_sf <- st_read("data/Massachusetts_Public_School_Districts")
district_sf <- st_read("data/schooldistricts")

# other shapefiles:
counties_sf <- st_read("data/counties") # shapefile by counties

zipcode_sf <- st_read("data/zipcodes_nt") # shapefile by zipcode
# from https://www.mass.gov/info-details/massgis-data-zip-codes-5-digit-from-here-navteq

town_sf <- st_read("data/kx-massachusetts-municipalities-SHP") %>% select(TOWN, geometry) # shapefile by boston towns

# reading data from https://profiles.doe.mass.edu/statereport/computersciencecourse.aspx
lowincome <- read_excel("data/lowincome.xlsx", skip = 1, col_names = TRUE)
totalstudents <- read_excel("data/totalstudentcount.xlsx", skip = 1, col_names = TRUE) %>% 
  filter(`District Name` != "State Totals") %>% 
  mutate(total_cs = as.numeric(gsub(",","",`All Grades`)),
         total = as.numeric(gsub(",","",`Total Students`))
         ) 

# data on school and its county
school_counties <- read_excel("data/counties_school.xlsx")

```
```{r}
# testing out shapefile
tm_shape(town_sf) + tm_polygons()
```

```{r}
lowincome <- read_excel("data/lowincome.xlsx", skip = 1, col_names = TRUE) # reset file

# clean data to fix data type errors and calculate ratio:
lowincome <- lowincome %>% mutate(low_income_cs = as.numeric(gsub(",","",`All Grades`)),
                                   low_income_total = as.numeric(gsub(",","",`Total Students`)),
                                   ratio_low_income = low_income_total / low_income_total) %>% 
  filter(`District Name` != "State Totals") %>% 
  select(-`All Grades`, -`Total Students`)

# join low income data with data on zipcode
lowincome_sf <- left_join(lowincome, schoolinfo, by=c("District Name" = "Org Name"))

# to match the other dataset
lowincome_sf$Town <- toupper(lowincome_sf$Town)

# case where zipcode does not have a school:
none_zipcode_sf <- anti_join(zipcode_sf, lowincome_sf, by=c("POSTCODE" = "Zip")) %>%
  st_as_sf()

# join zipcode with shapefile data
lowincome_sf <- full_join(lowincome_sf, town_sf, by=c("Town" = "TOWN")) %>% st_as_sf()

```
low_income_cs: students who are low income and is taking >=1 CS class
low_income_total: total number of low income students
total_cs: total number of students taking >=1 CS class
total: all students

ratio_low_income: low_income_cs / low_income_total
  This tells us how many low income students out of all low income students are taking CS classes.

ratio_low_total_cs: low_income_cs / total_cs
  This tells us how many of the students taking CS classes are from low income families.
  
ratio_cs_total: total_cs / total
  This tells us how many students are taking CS calsses
  
```{r}
# join low income data with cs literacy data:
low_total <- full_join(lowincome, totalstudents, by = 'District Code') %>% 
  select(`District Name.x`, `District Code`, low_income_cs, low_income_total, total_cs, total, ratio_low_income)

# calculate ratio:
low_total <- low_total %>%  mutate(ratio_low_total_cs = low_income_cs / total_cs)

# join school name and school zipcode
low_total_sf <- left_join(low_total, schoolinfo, by = c("District Name.x" = "Org Name"))
# join school zipcode and then school sf
low_total_sf <- inner_join(low_total_sf, zipcode_sf, by=c("Zip" = "POSTCODE")) %>% st_as_sf()

# test map visual
tm_shape(low_total_sf) +
  tm_polygons(col="ratio_low_total_cs", alpha=0.75)

```
```{r}
# manipulate data on total cs literacy
total_cs <- totalstudents %>% mutate(ratio_cs_total = total_cs / total)

# join data with address data:
total_cs_sf <- left_join(total_cs, schoolinfo, by = c("District Name" = "Org Name"))
total_cs_sf <- inner_join(total_cs_sf, zipcode_sf, by = c("Zip" = "POSTCODE")) %>% st_as_sf()

# test mapping out particpation across state:
tm_shape(none_zipcode_sf) +
  tm_polygons() +
tm_shape(total_cs_sf) + 
  tm_polygons(col="ratio_cs_total", 
              title="Ratio of students taking CS classes",
              palette="Blues") +
  tm_layout(main.title = "CS Literacy in Massachusetts by Zipcode")  

```

```{r}
# changing tmap settings:
# tmap_mode("view")
# tmap_mode("plot")

# ggplot(lowincome_sf) + geom_sf(aes(fill=ratio_low_income))
# 
# tm_shape(none_zipcode_sf) + 
#   tm_polygons() + 
# tm_shape(lowincome_sf) + 
#   tm_polygons(col="ratio_low_income")


```


```{r}
# crunching data to get all columns in one df:
everything <- low_total %>% mutate(ratio_cs_total = total_cs / total,
                                   ratio_low_total = low_income_total / total) %>% 
  mutate(charter = ifelse(grepl("Charter", `District Name.x`), 1, 0))

# join school name with its town
everything_sf <- left_join(everything, schoolinfo, by = c("District Name.x" = "Org Name"))
everything_sf$Town <- toupper(everything_sf$Town)
everything_sf <- full_join(everything_sf, town_sf, by = c("Town" = "TOWN"))

everything_sf2 <- everything_sf %>% group_by(Town) %>% summarize(count=n(), mean(ratio_cs_total))
everything_sf2 <- right_join(everything_sf2, town_sf, by = c("Town" = "TOWN")) %>% st_as_sf()

# everything_sf2$`mean(ratio_cs_total)` <- everything_sf2$`mean(ratio_cs_total)` %>%  replace_na(-1)

# scatterplot on ratio -- each point is a district
ggplot(everything, aes(x=ratio_low_total, y= ratio_cs_total)) +
  geom_point(aes(color=factor(charter)), alpha=0.75) +
  labs(x="Low Income Proportion \n(total low income students / total students)", 
       y="CS Literacy Proportion \n(total students taking CS courses / total students)",
       title="Relationship between Low Income and CS Course Participation",
       color = "") +
  scale_x_continuous(limits = c(0, 1), breaks = seq(0, 1, 0.1)) +
  scale_y_continuous(limits = c(0, 1), breaks = seq(0, 1, 0.1)) +
  scale_color_manual(labels = c("Not Chartered", "Chartered"), values = c("mediumpurple1", "limegreen"))

# plot just chartered 
ggplot(everything %>% filter(charter == 1), 
       aes(x=ratio_low_total, y= ratio_cs_total)) +
  geom_point()

# zoom in
ggplot(everything %>% filter(total < 15000)) + 
  geom_point(aes(x=total, y=low_income_cs))

```

```{r}
# test mapping mean ratios:
tm_shape(everything_sf2) + 
  tm_polygons(col="mean(ratio_cs_total)",
              title="Ratio of students taking CS classes",
              palette="Blues") +
  tm_layout(main.title = "Average CS Literacy in Massachusetts by Town")  

```

```{r}
# test plotting no data areas:
ggplot(none_zipcode_sf) + geom_sf()

tm_shape(none_zipcode_sf) + tm_polygons()

```

```{r}
# trying to join data with counties
everything_counties <- everything %>% mutate(first_name = word(`District Name.x`, 1))

school_counties <- school_counties %>% mutate(first_name = word(SchoolName, 1))

everything_counties <- left_join(everything_counties, school_counties, by = "first_name")

duplicated <- duplicated(everything_counties$`District Name.x`)

everything_counties <- everything_counties[!duplicated, ]

everything_counties <- everything_counties %>% drop_na()

everything_counties_sum <- everything_counties %>% 
  group_by(County) %>% 
  summarize(mean_li_ratio = mean(ratio_low_total),
            mean_cs_ratio = mean(ratio_cs_total))

```

```{r}

# sort the County column by mean_li_ratio in descending order
sorted_counties <- everything_counties_sum %>%
  filter(variable == "mean_li_ratio") %>%
  arrange(desc(ratio_val)) %>%
  pull(County)

# create a new factor variable with sorted counti
everything_counties_sum_pivoted <- everything_counties_sum %>% 
         pivot_longer(!County, names_to = "variable", values_to = "ratio_val")

everything_counties_sum_pivoted$County <- factor(
  everything_counties_sum_pivoted$County,
  levels = sorted_counties
)


# plot the bar chart with sorted counties
ggplot(everything_counties_sum_pivoted) +
  geom_bar(aes(x = County, y = ratio_val, fill = variable),
           stat = 'identity',
           position = 'dodge',
           width=0.8) +
  geom_hline(aes(yintercept = 0.4), stat = "hline", linetype=2) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(x="County", 
       y="Proportion",
       title="Average CS Literacy and Low Income propotions by Massachusetts County",
       fill = "Variable") +
  scale_fill_manual(labels = c("\nAverage CS Literacy Rate \n(CS students / total students)\n",
                               "\nAverage Low Income Proportion \n(low income students / total students)\n"), 
                    values = c("mediumpurple1", "orange"))

```

```{r}
# testing scatter on mean ratios (each point is a district)
ggplot(everything_counties_sum) + 
  geom_point(aes(x=mean_li_ratio, y=mean_cs_ratio)) 

```
```{r}
passing <- read_excel("data/passing.xlsx")

passing <- passing %>% mutate(low_income = ifelse(`Percentage of Low Income Student` >= 40,
                                     1, 0))

ggplot(passing %>% filter(`Number of NGO` < 200), aes(x=`Number of NGO`, y=`% Score 3-5`)) + 
  geom_point() +
  geom_text(data = passing %>% filter(`Number of NGO` < 200, 
                                      Town %in% c("Worcester", "Waltham", "Brookline", "Lexington", "Somerville", "Springfield", "Framingham")), 
            aes(label = Town),
            nudge_y = 3, nudge_x = -1) +
  labs(x="Number of NGOs", 
       y="Percent receiving passing score (AP Scores 3-5)",
       title="Number of NGOs and AP Test Performance by Towns in Massachusetts") 

ggplot(passing) +
  geom_point(aes(x=`Percentage of Low Income Student`, y=`% Score 3-5`))

```

Suggested stretch goal:

```{r}
ggplot(everything_counties %>% filter(County == "SUFFOLK"), aes(x=ratio_low_total, y= ratio_cs_total)) +
  geom_point(aes(color=factor(charter)), alpha=0.75) +
  labs(x="Low Income Proportion \n(total low income students / total students)", 
       y="CS Literacy Proportion \n(total students taking CS courses / total students)",
       title="Relationship between Low Income and CS Course Participation In Suffolk County",
       color = "") +
  scale_x_continuous(limits = c(0, 1), breaks = seq(0, 1, 0.1)) +
  scale_y_continuous(limits = c(0, 1), breaks = seq(0, 1, 0.1)) +
  scale_color_manual(labels = c("Not Chartered", "Chartered"), values = c("mediumpurple1", "limegreen"))


```


